---
title: "ECPR Data Imputation"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, error=FALSE, warning = FALSE)
```

## Introduction

This document performs imputation for select variables in the cleaned ECPR dataset.

## Load Data

```{r load_data}
library(tidyverse)
library(lubridate)

load("clean_data/data.RData")

cat("Loaded data:", nrow(data), "rows x", ncol(data), "columns\n")
```

## Imputation Rules

1. **n_shocks**: Impute 0 when NULL
2. **field_intubation**: Impute 0 when NULL
3. **lucas**: Impute 0 when NULL
4. **call_911_dt**: Impute as (hems_request - 11 minutes) when call_911_dt is NULL and hems_request is not NULL. If both are NULL, leave as NULL. Do not modify existing non-NULL values.

## Pre-Imputation Summary

```{r pre_imputation_summary}
cat("=== Pre-Imputation Missing Values ===\n\n")
cat("n_shocks:         ", sum(is.na(data$n_shocks)), "missing\n")
cat("field_intubation: ", sum(is.na(data$field_intubation)), "missing\n")
cat("lucas:            ", sum(is.na(data$lucas)), "missing\n")
cat("call_911_dt:      ", sum(is.na(data$call_911_dt)), "missing\n")
cat("hems_request:     ", sum(is.na(data$hems_request)), "missing\n")
```

## Perform Imputation

```{r imputation}
# Count how many call_911_dt values will be imputed from hems_request
n_call_911_imputed <- sum(is.na(data$call_911_dt) & !is.na(data$hems_request))

data_imp <- data %>%
  mutate(
    # Impute 0 for n_shocks, field_intubation, and lucas when NULL
    n_shocks = if_else(is.na(n_shocks), 0, n_shocks),
    field_intubation = if_else(is.na(field_intubation), 0L, field_intubation),
    lucas = if_else(is.na(lucas), 0L, lucas),

    # Impute call_911_dt as hems_request - 11 minutes when:
    #   - call_911_dt is NULL
    #   - hems_request is NOT NULL
    # Uses arrest_date combined with hems_request time
    # Note: format() is required to strip microseconds from hms objects for proper parsing
    call_911_dt = if_else(
      is.na(call_911_dt) & !is.na(hems_request),
      ymd_hms(paste(arrest_date, format(hems_request, "%H:%M:%S")), tz = "UTC") - minutes(11),
      call_911_dt
    )
  )

cat("call_911_dt values imputed from (hems_request - 11 min):", n_call_911_imputed, "\n")
```

## Post-Imputation Summary

```{r post_imputation_summary}
cat("=== Post-Imputation Missing Values ===\n\n")
cat("n_shocks:         ", sum(is.na(data_imp$n_shocks)), "missing\n")
cat("field_intubation: ", sum(is.na(data_imp$field_intubation)), "missing\n")
cat("lucas:            ", sum(is.na(data_imp$lucas)), "missing\n")
cat("call_911_dt:      ", sum(is.na(data_imp$call_911_dt)), "missing\n")
```

## Missing call_911_dt After Imputation

The following tables show the number of rows where `call_911_dt` remains NULL after imputation.

### By Cohort

```{r call_911_missing_by_cohort}
missing_by_cohort <- data_imp %>%
  group_by(cohort) %>%
  summarise(
    total_rows = n(),
    call_911_dt_missing = sum(is.na(call_911_dt)),
    pct_missing = round(call_911_dt_missing / total_rows * 100, 1)
  )

knitr::kable(missing_by_cohort,
             col.names = c("Cohort", "Total Rows", "call_911_dt Missing", "% Missing"))
```

### By ECMO Status (ecmo_yn)

```{r call_911_missing_by_ecmo}
missing_by_ecmo <- data_imp %>%
  mutate(ecmo_yn_label = case_when(
    ecmo_yn == 1 ~ "Yes",
    ecmo_yn == 0 ~ "No",
    TRUE ~ "Unknown"
  )) %>%
  group_by(ecmo_yn_label) %>%
  summarise(
    total_rows = n(),
    call_911_dt_missing = sum(is.na(call_911_dt)),
    pct_missing = round(call_911_dt_missing / total_rows * 100, 1)
  )

knitr::kable(missing_by_ecmo,
             col.names = c("ECMO Initiated", "Total Rows", "call_911_dt Missing", "% Missing"))
```

### Crosstab: Cohort x ECMO Status

```{r call_911_missing_crosstab}
missing_crosstab <- data_imp %>%
  mutate(ecmo_yn_label = case_when(
    ecmo_yn == 1 ~ "ECMO: Yes",
    ecmo_yn == 0 ~ "ECMO: No",
    TRUE ~ "ECMO: Unknown"
  )) %>%
  group_by(cohort, ecmo_yn_label) %>%
  summarise(
    total_rows = n(),
    call_911_dt_missing = sum(is.na(call_911_dt)),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = ecmo_yn_label,
    values_from = c(total_rows, call_911_dt_missing),
    values_fill = 0
  )

knitr::kable(missing_crosstab)
```

## EMS Hospital Arrival Time Imputation

### Validation Check

First, let's check the relationship between `ems_hospital_dt`, `hems_depart`, and `hems_flight_time` for rows where all three values are present.

```{r ems_hospital_validation}
# Check rows where all three fields are NOT missing
validation_data <- data_imp %>%
  filter(!is.na(ems_hospital_dt) & !is.na(hems_depart) & !is.na(hems_flight_time))

cat("Rows where all three fields (ems_hospital_dt, hems_depart, hems_flight_time) are present:",
    nrow(validation_data), "\n\n")

if (nrow(validation_data) > 0) {
  # Calculate the difference between ems_hospital_dt and hems_depart in minutes
  # and compare to hems_flight_time
  validation_check <- validation_data %>%
    mutate(
      # Create datetime for hems_depart using arrest_date
      # Note: format() strips microseconds from hms for proper parsing
      hems_depart_dt = ymd_hms(paste(arrest_date, format(hems_depart, "%H:%M:%S")), tz = "UTC"),
      # Calculate actual difference in minutes
      actual_diff_minutes = as.numeric(difftime(ems_hospital_dt, hems_depart_dt, units = "mins")),
      # Compare to listed flight time
      flight_time_diff = actual_diff_minutes - hems_flight_time
    ) %>%
    select(patient_id, ems_hospital_dt, hems_depart, hems_depart_dt, hems_flight_time,
           actual_diff_minutes, flight_time_diff)

  cat("=== Validation Summary ===\n")
  cat("Mean difference (actual - listed flight time):",
      round(mean(validation_check$flight_time_diff, na.rm = TRUE), 2), "minutes\n")
  cat("Median difference:",
      round(median(validation_check$flight_time_diff, na.rm = TRUE), 2), "minutes\n")
  cat("SD of difference:",
      round(sd(validation_check$flight_time_diff, na.rm = TRUE), 2), "minutes\n")
  cat("Range:", round(min(validation_check$flight_time_diff, na.rm = TRUE), 2), "to",
      round(max(validation_check$flight_time_diff, na.rm = TRUE), 2), "minutes\n\n")

  # Show detailed table
  cat("=== Detailed Comparison ===\n")
  knitr::kable(validation_check,
               col.names = c("Record ID", "ems_hospital_dt", "hems_depart", "hems_depart_dt",
                            "hems_flight_time (min)", "Actual Diff (min)", "Difference"))

  # Identify rows with negative actual_diff_minutes (impossible: arrival before departure)
  negative_diff_ids <- validation_check %>%
    filter(actual_diff_minutes < 0) %>%
    pull(patient_id)

  if (length(negative_diff_ids) > 0) {
    cat("\n=== Negative Time Difference Detected ===\n")
    cat("Setting ems_hospital_dt to NULL for", length(negative_diff_ids), "rows with negative actual diff\n")

    # Set ems_hospital_dt to NA for rows with negative difference
    data_imp <- data_imp %>%
      mutate(ems_hospital_dt = if_else(patient_id %in% negative_diff_ids, as.POSIXct(NA), ems_hospital_dt))
  }
} else {
  cat("No rows found with all three fields present for validation.\n")
}
```

### Pre-Imputation Status

```{r ems_hospital_pre_imputation}
cat("=== ems_hospital_dt Pre-Imputation Status ===\n\n")
cat("ems_hospital_dt missing:", sum(is.na(data_imp$ems_hospital_dt)), "\n")
cat("hems_depart missing:", sum(is.na(data_imp$hems_depart)), "\n")
cat("hems_flight_time missing:", sum(is.na(data_imp$hems_flight_time)), "\n\n")

# Count candidates for imputation
n_ems_impute_candidates <- sum(
  is.na(data_imp$ems_hospital_dt) &
  !is.na(data_imp$hems_depart) &
  !is.na(data_imp$hems_flight_time)
)
cat("Rows where ems_hospital_dt is missing but hems_depart AND hems_flight_time are present:",
    n_ems_impute_candidates, "\n")
```

### Perform ems_hospital_dt Imputation

```{r ems_hospital_imputation}
data_imp <- data_imp %>%
  mutate(
    # Impute ems_hospital_dt when:
    #   - ems_hospital_dt IS missing
    #   - hems_depart is NOT missing
    #   - hems_flight_time is NOT missing
    # Formula: ems_hospital_dt = hems_depart + hems_flight_time
    # Note: format() strips microseconds from hms for proper parsing
    ems_hospital_dt = if_else(
      is.na(ems_hospital_dt) & !is.na(hems_depart) & !is.na(hems_flight_time),
      ymd_hms(paste(arrest_date, format(hems_depart, "%H:%M:%S")), tz = "UTC") + minutes(hems_flight_time),
      ems_hospital_dt
    )
  )

cat("ems_hospital_dt values imputed:", n_ems_impute_candidates, "\n")
cat("ems_hospital_dt still missing after imputation:", sum(is.na(data_imp$ems_hospital_dt)), "\n")
```

## Calculate Times

### call_to_cann_hosp_time

Time from 9-1-1 call to ECMO cannulation hospital arrival (in minutes).

```{r call_to_cann_hosp_time}
data_imp <- data_imp %>%
  mutate(
    call_to_cann_hosp_time = as.numeric(difftime(ems_hospital_dt, call_911_dt, units = "mins"))
  )

cat("=== call_to_cann_hosp_time Summary ===\n")
cat("Non-missing values:", sum(!is.na(data_imp$call_to_cann_hosp_time)), "\n")
cat("Missing values:", sum(is.na(data_imp$call_to_cann_hosp_time)), "\n\n")

# Check for negative values (should not occur)
n_negative <- sum(data_imp$call_to_cann_hosp_time < 0, na.rm = TRUE)
if (n_negative > 0) {
  cat("WARNING:", n_negative, "rows have negative call_to_cann_hosp_time\n")
} else {
  cat("All values are positive (as expected)\n")
}

cat("\nDistribution (minutes):\n")
print(summary(data_imp$call_to_cann_hosp_time))
```

### hems_launch_time

Time from HEMS request to HEMS en route (in minutes).

```{r hems_launch_time}
data_imp <- data_imp %>%
  mutate(
    # hms objects store time as seconds since midnight
    # Subtract directly and convert to minutes
    hems_launch_time = as.numeric(hems_enroute - hems_request) / 60,
    # Handle midnight crossing: if negative, add 24 hours (1440 minutes)
    hems_launch_time = if_else(hems_launch_time < 0, hems_launch_time + 1440, hems_launch_time)
  )

cat("=== hems_launch_time Summary ===\n")
cat("Non-missing values:", sum(!is.na(data_imp$hems_launch_time)), "\n")
cat("Missing values:", sum(is.na(data_imp$hems_launch_time)), "\n\n")

cat("Distribution (minutes):\n")
print(summary(data_imp$hems_launch_time))

z<-data_imp %>% filter(abs(call_to_cann_hosp_time) > 400
                       ) %>%
  select(2,3,5,9:15,23:29,38,39)
```

## Save Imputed Data

```{r save_data}
save(data_imp, file = "clean_data/data_imputed.RData")
```
